---
title: "pep"
output: html_document
editor_options: 
  chunk_output_type: console
---

### 1.0 Header
```{r setup, include=FALSE}
#packrat::init("IIV_disability_PEP")
knitr::opts_chunk$set(echo = TRUE)
options(descr.plot = FALSE, scipen = 50, digits = 3)

# load some packages
library(here)
library(tidyverse)
library(haven)
library(descr)
library(magrittr)

# two central functions
center<-function(var){
  var=var-mean(var,na.rm=T)
}

vif.mer <- function (fit) {
    ## adapted from rms::vif
    
    v <- vcov(fit)
    nam <- names(fixef(fit))

    ## exclude intercepts
    ns <- sum(1 * (nam == "Intercept" | nam == "(Intercept)"))
    if (ns > 0) {
        v <- v[-(1:ns), -(1:ns), drop = FALSE]
        nam <- nam[-(1:ns)]
    }
    
    d <- diag(v)^0.5
    v <- diag(solve(v/(d %o% d)))
    names(v) <- nam
    v
}
```



### 1.1. Read & transform data
```{r}
pep1 <- read_sas("f2f_age_perinterval.sas7bdat")
pep2 <- read_sas("month12itemdis_f2fcovars21112018.sas7bdat")

# merge pep1 + pep2
pep <- left_join(pep2, pep1, by = c("StudyID", "interval")) %>%
       arrange(StudyID) 

# focus on terminal decline, i.e. on deceased respondents only
freq(pep$yearofdeath) # 17% of obs. missing year of death
freq(pep$timetodeath) # 20.5%
freq(pep$permrefuser) # 

# ID-level datset: how many respondents have died during follow-up (20yrs)
pep %>% 
  count(StudyID, timetodeath) %>%
  rename(num.int = n) %>% 
  arrange(StudyID) %>%
  mutate(dead = case_when(
    !is.na(timetodeath) ~ 1,
    is.na(timetodeath) ~ 0)) %>%
  count(dead) %>%
  mutate(freq = n / sum(n))



# Keep only those with valid timetodeath, aka deceased respondents
pep %<>%
  filter(!is.na(timetodeath)) 

# calculate time-to-death and keep only last 10 years of life as follow-up
pep %<>% 
  mutate(
    time.to.death = (intbloc - timetodeath)*-1,
    time.to.death.yrs = time.to.death / 12) %>%
  filter(time.to.death >= 0 & time.to.death <= 120)
hist(pep$time.to.death)


# recode categorical independent variables with simple coding
pep %<>%
  mutate(
    non.white = case_when(
      white == 1 ~ -0.5,
      white == 0 ~ 0.5),
    female = case_when(
      female == 1 ~ 0.5,
      female == 0 ~ -0.5),
    married = case_when(
      married == 1 ~ 0.5,
      married == 0 ~ -0.5),
    low.soc.supp = case_when(
      low_social_support == 1 ~ 0.5,
      low_social_support == 0 ~ -0.5),
    alone = case_when(
      LIVALNFU == 1 ~ 0.5, 
      LIVALNFU == 0 ~ -0.5),
    low.edu = case_when(
      edu_lt12 == 1 ~ 0.5, 
      edu_lt12 == 0 ~ -0.5),
    low.mmse = case_when(
      mmse_lt24 == 1 ~ 0.5,
      mmse_lt24 == 0 ~ -0.5), 
    frail = case_when(
      frail_d == 1 ~ 0.5,
      frail_d == 0 ~ -0.5)
    )

freq(pep$alone)
# Some subsetting of cases (to 657)
library(VIM)
# aggr_plot <- aggr(pep, col=c('navyblue','red'),
#                   numbers=TRUE, sortVars=TRUE,
#                   labels=names(pep),
#                   cex.axis=.7, gap=3,
#                   ylab=c("Histogram of missing data","Pattern"))

pep %<>% 
  filter(!is.na(female), 
    !is.na(time.to.death.yrs),
    !is.na(non.white),
    !is.na(married),
    !is.na(N_dis12),
    !is.na(chronconFU),
    !is.na(low.edu), 
    !is.na(low.soc.supp), 
    !is.na(low.mmse), 
    !is.na(frail),
    !is.na(alone))


# ID-level datset
pep.ID <- pep %>% 
  count(StudyID) %>%
  rename(num.int = n) %>% 
  filter(num.int >= 6) %>% # keep only resp. with 6 or more obs.
  arrange(StudyID) 

freq(pep.ID$num.int) 

# make new consistent id variable
pep.ID %<>%
  mutate(id = rownames(pep.ID),
         id = as.numeric(id)) 

# bind back and modify pep-observations-level df
pep <- right_join(pep, pep.ID[,c("StudyID", "id", "num.int")], 
                  by = "StudyID")


## some re-naming and transforming ...
pep %<>% 
  mutate( edu.cont.c = center(edu_cont), 
    num.int.c = center(num.int),
    age.b = age_b, 
    age.b.70 = age_b - 70, # center at age 70
    age = age_fu,
    age.c = center(age),
    time.to.death.yrs.c = center(time.to.death.yrs), 
    morbidity.c = center(chronconFU))


# calculate age at death (=predictor for time-to-death analysis)
pep.x <- pep %>%  
  group_by(id) %>%
  summarise(age.dead = max(age))
pep <- left_join(pep, pep.x, by = "id")
pep %<>%
  mutate(age.dead.c = center(age.dead))
hist(pep$age.dead)

    
# make line plots for the first 12 individuals
library(ggformula)
pep.100 <- subset(pep, id <= 12)
ggplot(pep.100, aes(x=time.to.death, y=N_dis12)) +
    geom_smooth(se=T) + 
    geom_spline(all.knots=F, nknots=3, color="red", size=1) + 
  geom_line(size=0.25) +
  facet_wrap(~id, nrow=3) +
  scale_y_continuous("Disability", breaks = seq(0,12,2),
                     limits=c(0,12)) +
  scale_x_reverse("Months to death", breaks = seq(0,120,24),
                  limits=c(120,0)) +
  theme_bw(12) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(), 
        legend.position = "top") +
  ggsave("app_fig1_raw_loess.png", 
         height = 6, width = 9, dpi=600)

?smooth.spline()

```


### 1.2. Descriptives
```{r}
### a couple of descriptives
pep %>%  
  count(female) %>%
  mutate(prop = prop.table(n))

# N_dis12: Mean, median + SD by sex
pep %>%  
  group_by(female) %>%
  summarise(mean = mean(N_dis12, na.rm=T), 
         median = median(N_dis12, na.rm=T),
         sd = sd(N_dis12, na.rm=T),
         miss.n = sum(is.na(N_dis12)))

# Baseline Age: Mean, min, max by sex
pep %>%  
  group_by(female) %>%
  summarise(mean=mean(age_b), 
         min=min(age_b),
         max=max(age_b))
hist(pep$age_b)



# descriptives for number of monhtly interviews
pep.ID %>%  
  group_by(female) %>%
  summarise(mean.num.month.int=mean(num.int),
            sd.num.month.int=sd(num.int),
         median.num.month.int=median(num.int),
         min.num.month.int=min(num.int),
         max.num.month.int=max(num.int))
hist(pep.ID$num.int)

```


# 1.3. Calculate iSD and mssd (and bind back to pep.ID)
```{r}
library(psych)
i.SD <- pep %>%  
  group_by(StudyID) %>%
  summarise(imean = mean(N_dis12, na.rm=T),
            iSD = sd(N_dis12, na.rm=T),
            rmssd = rmssd(N_dis12, na.rm=T),
            min = min(N_dis12, na.rm=T), 
            max = max(N_dis12, na.rm=T)) 
pep.ID <- left_join(pep.ID, 
                      i.SD[,c("StudyID", "iSD", "rmssd", "imean")], 
                      by = "StudyID")

pep.ID %>%  
  summarise(mean.iSD = mean(iSD, na.rm=T),
            median.iSD= median(iSD, na.rm=T),
            mean.mssd = mean(rmssd, na.rm=T),
            median.mssd= median(rmssd, na.rm=T))
pep <- left_join(pep, pep.ID[,c("StudyID", "iSD", "rmssd", "imean")],
                 by = "StudyID")

# calculate correlation between iMEAN and iSD
cor(pep.ID$imean, pep.ID$iSD, use="complete.obs")
cor(pep.ID$imean, pep.ID$rmssd, use="complete.obs")


# calculate baseline N_dis12
pep.b <- subset(pep, intbloc == 1)
pep.b$N_dis12.b <- pep.b$N_dis12
pep.b <- pep.b[,c("StudyID", "N_dis12.b")]
pep <- left_join(pep, pep.b, by = "StudyID")
check <- pep[,c("StudyID", "intbloc", "N_dis12.b", "N_dis12")]
freq(pep$N_dis12.b) # ~35% of respondents had 1+ disability at baseline
```

#  make 24months dataset (comparable to df in Lin 2015)
```{r}
freq(pep$time.to.death)
pep24m <- subset(pep, time.to.death == 0 | time.to.death == 24 | 
                   time.to.death == 48 |
                   time.to.death == 72 | time.to.death == 96 | 
                   time.to.death == 120)

# make wide df by StudyID
pep24m.w.ID <- pep24m %>% 
  count(id, female, age.b) %>%
  arrange(id) %>%
  rename(num.int = n)

# descriptives for number of monhtly interviews
pep24m.w.ID %>%  
  group_by(female) %>%
  summarise(mean.num.biann.int=mean(num.int),
            sd.num.biann.int=sd(num.int),
         median.num.biann.int=median(num.int),
         min.num.biann.int=min(num.int),
         max.num.biann.int=max(num.int))

# calculate iSD and mssd (and bind back to pep.ID)
library(psych)
i.SD24 <- pep24m %>%  
  group_by(id) %>%
  summarise(imean24 = mean(N_dis12, na.rm=T),
            iSD24 = sd(N_dis12, na.rm=T),
            rmssd24 = rmssd(N_dis12, na.rm=T),
            min24 = min(N_dis12, na.rm=T), 
            max24 = max(N_dis12, na.rm=T)) 
pep.ID <- left_join(pep.ID, i.SD24, by = "StudyID")

hist(pep.ID$iSD)
hist(pep.ID$iSD24)
hist(pep.ID$rmssd)
hist(pep.ID$rmssd24)

pep.ID %>%  
  group_by(female) %>%
  summarise(median.iSD = median(iSD, na.rm=T),
            median.iSD24= median(iSD24, na.rm=T),
            median.rmssd = median(rmssd, na.rm=T),
            median.rmssd24= median(rmssd24, na.rm=T))
```


### 1.4. Unconditional model (lme4)
```{r}
library(lme4)
### STEP 1: means-model
m1.lmer <- lmer(N_dis12 ~  time.to.death.yrs.c +  
              num.int.c + 
             (1 + time.to.death.yrs.c | id), 
           pep, REML=F)
summary(m1.lmer)

m2.lmer <- lmer(N_dis12 ~  time.to.death.yrs.c + I(time.to.death.yrs.c^2)  + 
              num.int.c + 
             (1 + time.to.death.yrs.c | id), 
           pep, REML=F)
summary(m2.lmer)
anova(m1.lmer, m2.lmer)

m3.lmer <- lmer(N_dis12 ~  time.to.death.yrs.c + 
                  I(time.to.death.yrs.c^2)  + I(time.to.death.yrs.c^3) +
              num.int.c + 
             (1 + time.to.death.yrs.c | id), 
           pep, REML=F)
summary(m3.lmer)
anova(m2.lmer, m3.lmer)

m4.lmer <- lmer(N_dis12 ~  time.to.death.yrs.c + 
                  I(time.to.death.yrs.c^2)  + I(time.to.death.yrs.c^3) +
                  I(time.to.death.yrs.c^4) +
              num.int.c + 
             (1 + time.to.death.yrs.c | id), 
           pep, REML=F)
summary(m4.lmer)
anova(m3.lmer, m4.lmer)
library(MuMIn)
round(Weights(AICc(m1.lmer, m2.lmer, m3.lmer, m4.lmer)), 3) # keep 
# quartic model, because little substantial change after that but scaling
# problems due to very small effects from quintic onwards. 
# Also, splines show highly similar fit. 


m5.lmer <- lmer(N_dis12 ~  time.to.death.yrs.c + 
                  I(time.to.death.yrs.c^2)  + I(time.to.death.yrs.c^3) +
                  I(time.to.death.yrs.c^4) + I(time.to.death.yrs.c^5) +
              num.int.c + 
             (1 + time.to.death.yrs.c | id), 
           pep, REML=F)
summary(m5.lmer)
anova(m4.lmer, m5.lmer)

library(splines)
# with 3 knots
m6.lmer <- lmer(N_dis12 ~  ns(time.to.death.yrs.c, df=3) +
              num.int.c + 
             (1 + time.to.death.yrs.c | id), 
           pep, REML=F)
summary(m6.lmer)
anova(m5.lmer, m6.lmer)
round(Weights(AICc(m4.lmer, m5.lmer, m6.lmer)), 3) 

m7.lmer <- lmer(N_dis12 ~  time.to.death.yrs.c + 
                  I(time.to.death.yrs.c^2)  + I(time.to.death.yrs.c^3) +
                  I(time.to.death.yrs.c^4) + 
             (1 + time.to.death.yrs.c + I(time.to.death.yrs.c^2)| id), 
           pep, REML=F)
summary(m7.lmer)
anova(m5.lmer, m7.lmer)

library(splines)
# with 3 knots
m8.lmer <- lmer(N_dis12 ~  ns(time.to.death.yrs.c, df=2) +
              num.int.c + 
             (1 + ns(time.to.death.yrs.c, df=2) | id), 
           pep, REML=F)
summary(m8.lmer)
anova(m6.lmer, m8.lmer)
coef(m8.lmer)
```

#### 1.4.1 Predictions for model comparison
```{r}
library(merTools)
## plot results
# linear
newData <- with(pep, expand.grid(time.to.death.yrs.c = 
                                   unique(time.to.death.yrs.c), 
                                 num.int.c = mean(num.int.c),
                                 id = 20))
PI <- predictInterval(merMod = m1.lmer, newdata = newData,
                      level = 0.95, n.sims = 1000, which = "fixed",
                      stat = "median", type="linear.prediction",
                      include.resid.var = F)
newData <- cbind(newData, PI)
newData$time.to.death.yrs <- 
  newData$time.to.death.yrs.c  + mean(pep$time.to.death.yrs)

library(ggplot2)
fig.linear <- ggplot(data = newData, 
       aes(x = time.to.death.yrs, y = fit)) +
  geom_line(aes(y = upr, x = time.to.death.yrs),  
            size = 0.3, alpha=0.5, linetype=2) + 
  geom_line(aes(y = lwr, x = time.to.death.yrs), 
            size = 0.3, alpha=0.5, linetype=2) +
  geom_line(aes(y = fit, x = time.to.death.yrs), size = 1) + 
  scale_y_continuous("Pred. disability", breaks = seq(0,12,1)) +
  scale_x_reverse("Time to death (yrs)", breaks=seq(0,10,1)) +
  geom_hline(yintercept = 12, linetype=2, color="red")  +
  theme_bw(10)

# quadratic
newData <- with(pep, expand.grid(time.to.death.yrs.c = 
                                   unique(time.to.death.yrs.c), 
                                 num.int.c = mean(num.int.c),
                                 id = 20))
PI <- predictInterval(merMod = m2.lmer, newdata = newData,
                      level = 0.95, n.sims = 1000, which = "fixed",
                      stat = "median", type="linear.prediction",
                      include.resid.var = F)
newData <- cbind(newData, PI)
newData$time.to.death.yrs <- 
  newData$time.to.death.yrs.c  + mean(pep$time.to.death.yrs)

fig.quadr <- ggplot(data = newData, 
       aes(x = time.to.death.yrs, y = fit)) +
  geom_line(aes(y = upr, x = time.to.death.yrs),  
            size = 0.3, alpha=0.5, linetype=2) + 
  geom_line(aes(y = lwr, x = time.to.death.yrs), 
            size = 0.3, alpha=0.5, linetype=2) +
  geom_line(aes(y = fit, x = time.to.death.yrs), size = 1) + 
  scale_y_continuous("Pred. disability", breaks = seq(0,12,1)) +
  scale_x_reverse("Time to death (yrs)", breaks=seq(0,10,1), 
                     limits=c(10,0)) +
  geom_hline(yintercept = 12, linetype=2, color="red")  +
  theme_bw(10)

# cubic
newData <- with(pep, expand.grid(time.to.death.yrs.c = 
                                   unique(time.to.death.yrs.c), 
                                 num.int.c = mean(num.int.c),
                                 id = 20))
PI <- predictInterval(merMod = m3.lmer, newdata = newData,
                      level = 0.95, n.sims = 1000, which = "fixed",
                      stat = "median", type="linear.prediction",
                      include.resid.var = F)
newData <- cbind(newData, PI)
newData$time.to.death.yrs <- 
  newData$time.to.death.yrs.c  + mean(pep$time.to.death.yrs)

fig.cubic <- ggplot(data = newData, 
       aes(x = time.to.death.yrs, y = fit)) +
  geom_line(aes(y = upr, x = time.to.death.yrs),  
            size = 0.3, alpha=0.5, linetype=2) + 
  geom_line(aes(y = lwr, x = time.to.death.yrs), 
            size = 0.3, alpha=0.5, linetype=2) +
  geom_line(aes(y = fit, x = time.to.death.yrs), size = 1) + 
  scale_y_continuous("Pred. disability", breaks = seq(0,12,1)) +
  scale_x_reverse("Time to death (yrs)", breaks=seq(0,10,1), 
                     limits=c(10,0)) +
  geom_hline(yintercept = 12, linetype=2, color="red")  +
  theme_bw(10)

# quartic
newData <- with(pep, expand.grid(time.to.death.yrs.c = 
                                   unique(time.to.death.yrs.c), 
                                 num.int.c = mean(num.int.c),
                                 id = 20))
PI <- predictInterval(merMod = m4.lmer, newdata = newData,
                      level = 0.95, n.sims = 1000, which = "fixed",
                      stat = "median", type="linear.prediction",
                      include.resid.var = F)
newData <- cbind(newData, PI)
newData$time.to.death.yrs <- 
  newData$time.to.death.yrs.c  + mean(pep$time.to.death.yrs)

fig.quartic <- ggplot(data = newData, 
       aes(x = time.to.death.yrs, y = fit)) +
  geom_line(aes(y = upr, x = time.to.death.yrs),  
            size = 0.3, alpha=0.5, linetype=2) + 
  geom_line(aes(y = lwr, x = time.to.death.yrs), 
            size = 0.3, alpha=0.5, linetype=2) +
  geom_line(aes(y = fit, x = time.to.death.yrs), size = 1) + 
  scale_y_continuous("Pred. disability", breaks = seq(0,12,1)) +
  scale_x_reverse("Time to death (yrs)", breaks=seq(0,10,1), 
                     limits=c(10,0)) +
  geom_hline(yintercept = 12, linetype=2, color="red")  +
  theme_bw(10)

# quintic
newData <- with(pep, expand.grid(time.to.death.yrs.c = 
                                   unique(time.to.death.yrs.c), 
                                 num.int.c = mean(num.int.c),
                                 id = 20))
PI <- predictInterval(merMod = m5.lmer, newdata = newData,
                      level = 0.95, n.sims = 1000, which = "fixed",
                      stat = "median", type="linear.prediction",
                      include.resid.var = F)
newData <- cbind(newData, PI)
newData$time.to.death.yrs <- 
  newData$time.to.death.yrs.c  + mean(pep$time.to.death.yrs)

fig.quintic <- ggplot(data = newData, 
       aes(x = time.to.death.yrs, y = fit)) +
  geom_line(aes(y = upr, x = time.to.death.yrs),  
            size = 0.3, alpha=0.5, linetype=2) + 
  geom_line(aes(y = lwr, x = time.to.death.yrs), 
            size = 0.3, alpha=0.5, linetype=2) +
  geom_line(aes(y = fit, x = time.to.death.yrs), size = 1) + 
  scale_y_continuous("Pred. disability", breaks = seq(0,12,1)) +
  scale_x_reverse("Time to death (yrs)", breaks=seq(0,10,1), 
                     limits=c(10,0)) +
  geom_hline(yintercept = 12, linetype=2, color="red")  +
  theme_bw(10)

# spline
newData <- with(pep, expand.grid(time.to.death.yrs.c = 
                                   unique(time.to.death.yrs.c), 
                                 num.int.c = mean(num.int.c),
                                 id = 20))
PI <- predictInterval(merMod = m6.lmer, newdata = newData,
                      level = 0.95, n.sims = 1000, which = "fixed",
                      stat = "median", type="linear.prediction",
                      include.resid.var = F)
newData <- cbind(newData, PI)
newData$time.to.death.yrs <- 
  newData$time.to.death.yrs.c  + mean(pep$time.to.death.yrs)

fig.spline <- ggplot(data = newData, 
       aes(x = time.to.death.yrs, y = fit)) +
  geom_line(aes(y = upr, x = time.to.death.yrs),  
            size = 0.3, alpha=0.5, linetype=2) + 
  geom_line(aes(y = lwr, x = time.to.death.yrs), 
            size = 0.3, alpha=0.5, linetype=2) +
  geom_line(aes(y = fit, x = time.to.death.yrs), size = 1) + 
  scale_y_continuous("Pred. disability", breaks = seq(0,12,1)) +
  scale_x_reverse("Time to death (yrs)", breaks=seq(0,10,1), 
                     limits=c(10,0)) +
  geom_hline(yintercept = 12, linetype=2, color="red")  +
  theme_bw(10)

# quadratic slope (w. polynomial)
newData <- with(pep, expand.grid(time.to.death.yrs.c = 
                                   unique(time.to.death.yrs.c), 
                                 num.int.c = mean(num.int.c),
                                 id = 20))
PI <- predictInterval(merMod = m7.lmer, newdata = newData,
                      level = 0.95, n.sims = 1000, which = "fixed",
                      stat = "median", type="linear.prediction",
                      include.resid.var = F)
newData <- cbind(newData, PI)
newData$time.to.death.yrs <- 
  newData$time.to.death.yrs.c  + mean(pep$time.to.death.yrs)

fig.quadr.slope <- ggplot(data = newData, 
       aes(x = time.to.death.yrs, y = fit)) +
  geom_line(aes(y = upr, x = time.to.death.yrs),  
            size = 0.3, alpha=0.5, linetype=2) + 
  geom_line(aes(y = lwr, x = time.to.death.yrs), 
            size = 0.3, alpha=0.5, linetype=2) +
  geom_line(aes(y = fit, x = time.to.death.yrs), size = 1) + 
  scale_y_continuous("Pred. disability", breaks = seq(0,12,1)) +
  scale_x_reverse("Time to death (yrs)", breaks=seq(0,10,1)) +
  geom_hline(yintercept = 12, linetype=2, color="red")  +
  theme_bw(10)

# slope with spline
newData <- with(pep, expand.grid(time.to.death.yrs.c = 
                                   unique(time.to.death.yrs.c), 
                                 num.int.c = mean(num.int.c),
                                 id = unique(id)))
PI <- predictInterval(merMod = m8.lmer, newdata = newData,
                      level = 0.95, n.sims = 1000, which = "all",
                      stat = "median", type="linear.prediction",
                      include.resid.var = F)
newData <- cbind(newData, PI)
newData$time.to.death.yrs <- 
  newData$time.to.death.yrs.c  + mean(pep$time.to.death.yrs)

fig.spline.slope <- ggplot(data = newData, 
       aes(x = time.to.death.yrs, y = fit)) +
  geom_line(aes(y = upr, x = time.to.death.yrs),  
            size = 0.3, alpha=0.5, linetype=2) + 
  geom_line(aes(y = lwr, x = time.to.death.yrs), 
            size = 0.3, alpha=0.5, linetype=2) +
  geom_line(aes(y = fit, x = time.to.death.yrs), size = 1) + 
  scale_y_continuous("Pred. disability", breaks = seq(0,12,1)) +
  scale_x_reverse("Time to death (yrs)", breaks=seq(0,10,1)) +
  geom_hline(yintercept = 12, linetype=2, color="red")  +
  theme_bw(10)

library(cowplot)
plot_grid(fig.linear, fig.quadr, fig.cubic, fig.quartic, 
          fig.quintic, fig.spline, fig.quadr.slope, fig.spline.slope,
          ncol = 4, align = 'h', labels="auto") +
  ggsave(filename="fig_fits.png", units="cm", width=20, height=12, dpi=600)
```


#### Comparing plots with and without residual variation
```{r}
newData <- with(pep, expand.grid(time.to.death.yrs.c = 
                                   unique(time.to.death.yrs.c), 
                                 num.int.c = mean(num.int.c),
                                 id = 20))
PI <- predictInterval(merMod = m4.lmer, newdata = newData,
                      level = 0.95, n.sims = 1000, which = "full",
                      stat = "median", type="linear.prediction",
                      include.resid.var = F)
newData <- cbind(newData, PI)
newData$time.to.death.yrs <- 
  newData$time.to.death.yrs.c  + mean(pep$time.to.death.yrs)

library(ggplot2)
fig.quartic1 <- ggplot(data = newData, 
       aes(x = time.to.death.yrs, y = fit)) +
  geom_line(aes(y = upr, x = time.to.death.yrs),  
            size = 0.3, alpha=0.5, linetype=2) + 
  geom_line(aes(y = lwr, x = time.to.death.yrs), 
            size = 0.3, alpha=0.5, linetype=2) +
  geom_line(aes(y = fit, x = time.to.death.yrs), size = 1) + 
  scale_y_continuous("Pred. disability", breaks = seq(0,12,1)) +
  scale_x_reverse("Time to death (yrs)", breaks=seq(0,10,1), 
                     limits=c(10,0)) +
  geom_hline(yintercept = 12, linetype=2, color="red")  +
  theme_bw(10)

newData <- with(pep, expand.grid(time.to.death.yrs.c = 
                                   unique(time.to.death.yrs.c), 
                                 num.int.c = mean(num.int.c),
                                 id = 20))
PI <- predictInterval(merMod = m4.lmer, newdata = newData,
                      level = 0.95, n.sims = 1000, which = "fixed",
                      stat = "median", type="linear.prediction",
                      include.resid.var = T)
newData <- cbind(newData, PI)
newData$time.to.death.yrs <- 
  newData$time.to.death.yrs.c  + mean(pep$time.to.death.yrs)

library(ggplot2)
fig.quartic2 <- ggplot(data = newData, 
       aes(x = time.to.death.yrs, y = fit)) +
  geom_line(aes(y = upr, x = time.to.death.yrs),  
            size = 0.3, alpha=0.5, linetype=2) + 
  geom_line(aes(y = lwr, x = time.to.death.yrs), 
            size = 0.3, alpha=0.5, linetype=2) +
  geom_line(aes(y = fit, x = time.to.death.yrs), size = 1) + 
  scale_y_continuous("Pred. disability", breaks = seq(0,12,1)) +
  scale_x_reverse("Time to death (yrs)", breaks=seq(0,10,1), 
                     limits=c(10,0)) +
  geom_hline(yintercept = 12, linetype=2, color="red")  +
  theme_bw(10)

library(cowplot)
plot_grid(fig.quartic1, fig.quartic2,
          ncol = 1, align = 'h', labels="auto") +
  ggsave(filename="fig_residuals.png", units="cm", width=8, height=8, dpi=600)
```


### 1.5. Conditional model (lme4)
```{r}
# adding covariates
m9.lmer <- lmer(N_dis12 ~  time.to.death.yrs.c +
                  I(time.to.death.yrs.c^2)  + 
                  I(time.to.death.yrs.c^3) +
                  I(time.to.death.yrs.c^4) +
             age.dead.c + female +  low.edu + non.white + 
                morbidity.c + frail + low.mmse + 
                num.int.c +
             (1 + time.to.death.yrs.c | id), 
           pep, REML=F)
summary(m9.lmer)
library(MuMIn)
r.squaredGLMM(m9.lmer)
vif.mer(m9.lmer)
qqnorm(resid(m9.lmer))
qqline(resid(m9.lmer))

library(stargazer)
stargazer(m6.lmer, title=c("Disability trajectories"), 
          ci=T, single.row=T, align=T, digits=2, 
          intercept.top=T, intercept.bottom = F, no.space=T)


```



### 1.6. Modell disability IIV (lme4)
```{r}
pep$res <- resid(m9.lmer) 
hist(pep$res)
pep$res.a <- abs(pep$res) #take absolute values


# calculate geometric mean (acts as baseline for interpretation of %
# effects from logged outcome)
gm_mean = function(x, na.rm=TRUE){
  exp(sum(log(x[x > 0]), na.rm=na.rm) / length(x))
}
gm_mean(pep$res.a)
pep$res.l <- log(pep$res.a) # log due to skew
par(mfrow=c(1,3))
hist(pep$res)
hist(pep$res.a)
hist(pep$res.l)

# Using linear model and log transformation beforehand
m10.lmer <- lmer(res.l ~ time.to.death.yrs.c +
             (1 | id), 
           pep, REML=F)
summary(m10.lmer)

m11.lmer <- lmer(res.l ~ time.to.death.yrs.c + I(time.to.death.yrs.c^2) + 
             (1 | id), 
           pep, REML=F)
summary(m11.lmer)
anova(m10.lmer, m11.lmer)

m12.lmer <- lmer(res.l ~ time.to.death.yrs.c + I(time.to.death.yrs.c^2) + 
             I(time.to.death.yrs.c^3) + 
             (1 | id), 
           pep, REML=F)
summary(m12.lmer)
anova(m11.lmer, m12.lmer) # no improvement: quadratic effect it is!

# INterpretation: LOG SCALE!
install.packages("broom.mixed")
library(broom.mixed)
tidy(m8.lmer,conf.int=TRUE,exponentiate=TRUE,effects="fixed")

# plot model 8
newdata1 <- with(pep, expand.grid(time.to.death.yrs.c =
                                    unique(time.to.death.yrs.c),
                                      id = 1))
library(merTools)
PI <- predictInterval(merMod = m11.lmer, newdata = newdata1,
                      level = 0.95, n.sims = 1000, which = "fixed",
                      stat = "median", type="linear.prediction",
                      include.resid.var = F)
newdata1 <- cbind(newdata1, PI)
newdata1$time.to.death.yrs <- newdata1$time.to.death.yrs.c +
  mean(pep$time.to.death.yrs)
library(ggplot2)
ggplot(data = newdata1, 
               aes(x = time.to.death.yrs, y = fit)) +
  geom_line(aes(y = upr, x = time.to.death.yrs),  
            size = 0.3, alpha=0.5, linetype=2) + 
  geom_line(aes(y = lwr, x = time.to.death.yrs), 
            size = 0.3, alpha=0.5, linetype=2) +
  geom_line(aes(y = fit, x = time.to.death.yrs), size = 1) + 
  scale_y_continuous("log(Disability IIV)") +
  scale_x_reverse("Time to death (yrs)", breaks=seq(0,10,1),
                  limits=c(10,0))

exp(-0.4)
exp(0.2)
```

```{r}
# Adding covariates
m13.lmer <- lmer(res.l ~ imean + time.to.death.yrs.c +
                  I(time.to.death.yrs.c^2) + 
             age.dead + female +  low.edu + non.white + 
                num.int.c +
             (1 | id), 
           pep, REML=F)
summary(m13.lmer)
library(MuMIn)
r.squaredGLMM(m13.lmer)
confint(m13.lmer, level=0.95, method="Wald")

names(pep)

```

### 1.7. Modell autocorrelation in disability (lme4)
```{r}
pep.lag <- pep %>% # make lag variable of resid
    group_by(id) %>% 
  mutate(lag1=lag(res)) %>%
  filter(!is.na(lag1))

## difference between t_1 und t_-1 = Measure for autocorrelation (and iSD # rspectively their interaction!), see Wang et al. 2013, p.6-5. 
# Only difference to MSSD: here, we are using it as an observation-level
# figure (just the difference, no squaring, no summing) and not as the sum of squared differences (=MSSD). Wobei, Squaring hat den Sinn das größere Abweichungungen stärker gewichtet werden ...
pep.lag$autcorr.diff <- pep.lag$res - pep.lag$lag1
pep.lag$autcorr.diff.abs <- abs(pep.lag$autcorr.diff)
pep.lag$autcorr.diff.abs.log <- log(pep.lag$autcorr.diff.abs)

# descriptives
par(mfrow=c(1,3))
hist(pep.lag$autcorr.diff)
hist(pep.lag$autcorr.diff.abs)
hist(pep.lag$autcorr.diff.abs.log)


check <- pep.lag[,c("id", "intbloc", "N_dis12", "res", 
                    "lag1", "autcorr.diff", "autcorr.diff.abs")]

# Adding covariates
mx.lmer <- lmer(autcorr.diff.abs.log ~ imean + time.to.death.yrs.c +
                  I(time.to.death.yrs.c^2) + 
             age.dead + female +  low.edu + non.white + 
                num.int.c +
             (1 | id), 
           pep.lag, REML=F)
summary(mx.lmer)
library(MuMIn)
r.squaredGLMM(mx.lmer)
confint(mx.lmer, level=0.95, method="Wald")
```


# Alternativ: model residuals with quasibinomial
```{r}
#negative binomial from GLMMadaptive
library(GLMMadaptive)
m8.alt <- mixed_model(res.a ~   time.to.death.yrs.c +
                            I(time.to.death.yrs.c^2) + 
                            age.b.70 + 
                            num.int.c, 
                         random = ~ 1| id, 
                         family=negative.binomial(), data = pep)
summary(m8.alt) 
# need to plot it ...

m10.alt <- mixed_model(res.a ~   imean + time.to.death.yrs.c +
                            I(time.to.death.yrs.c^2) + 
                            age.b.70 + female +  low.edu + non.white + 
                            married + low.soc.supp +
                            morbidity.c +  low.mmse + 
                            num.int.c, 
                         random = ~ 1| id, 
                         family=negative.binomial(), data = pep)
summary(m10.alt) 
```

### 1.5. Export summary tables
```{r}
library(stargazer)
stargazer(m1.lmer, ci=T, single.row=T)
```


### 1.5. Model data with brms
```{r}
library(brms)
m1.brms <- brm(N_dis12 ~  time.to.death.yrs.c + 
                  I(time.to.death.yrs.c^2)  + 
                  I(time.to.death.yrs.c^3) +
                  I(time.to.death.yrs.c^4) +
                  age.b.70 + female +  low.edu + non.white + 
                  married + low.soc.supp +
                  morbidity.c + low.mmse + 
                  num.int.c +
                  (1 + time.to.death.yrs.c|ID|id), 
                  family = gaussian(), 
                  chains=4, warmup=2000, iter=12000,
                  data = pep, core=4, thin=2, 
                  control = list(adapt_delta = 0.95))
summary(m1.brms)

m2.brms <- brm(bf(N_dis12 ~  time.to.death.yrs.c + 
                  I(time.to.death.yrs.c^2)  + 
                  I(time.to.death.yrs.c^3) +
                  I(time.to.death.yrs.c^4) +
                  age.b.70 + female +  low.edu + non.white + 
                  married + low.soc.supp +
                  morbidity.c + low.mmse + 
                  num.int.c +
                  (1 + time.to.death.yrs.c|ID|id), 
                  
                  sigma ~  1 + (1|ID|id)),
             
                  family = gaussian(), 
                  chains=4, warmup=2000, iter=12000,
                  data = pep, core=4, thin=2, 
                  control = list(adapt_delta = 0.95))
summary(m2.brms)
saveRDS(m2.brms, file="m2_brms.rds")
plot(m2.brms)



m3.brms <- brm(bf(N_dis12 ~  time.to.death.yrs.c + 
                  I(time.to.death.yrs.c^2)  + 
                  I(time.to.death.yrs.c^3) +
                  I(time.to.death.yrs.c^4) +
                  age.dead.c + female +  low.edu + non.white + 
                  married + low.soc.supp +
                  morbidity.c + low.mmse + 
                  num.int.c +
                  (1 + time.to.death.yrs.c|ID|id), 
                  
                  sigma ~  time.to.death.yrs.c + 
                  I(time.to.death.yrs.c^2)  + 
                  age.dead.c + female +  low.edu + non.white + 
                  num.int.c +
                  (1|ID|id)),
             
                  family = gaussian(), 
                  chains=4, warmup=2000, iter=12000,
                  data = pep, core=4, thin=2, 
                  control = list(adapt_delta = 0.99))
summary(m3.brms)
plot(m3.brms)
saveRDS(m3.brms, file="m3_brms.rds")
```


```{r}
### CORE PROBLEM: ADLS and linear model dont go well together ...
# other distributions?
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
pep %>%  
  summarise(mean=mean(N_dis12),
            var=var(N_dis12),
            mode=Mode(N_dis12),
            min=min(N_dis12), 
            max=max(N_dis12))
hist(pep$N_dis12)
freq(pep$N_dis12)
13.3/3.23
# i.e.: clearly skewed, low mean, high variance (=overdispersed). Only positive numbers and integers. This rules linear model and ... possion model out. 
hist(log(pep$N_dis12)) # log transformation does not normalise things (enough) 
# either, thus, a log transformation and a linear model wont do. SEE: 
#https://www.theanalysisfactor.com/count-data-considered-continuous/

# i.e. what remains: poisson, negative binomial, binomial, beta-binomial ...
# beta-binomial machen wohl bbmle als auch GLMMadaptive (unter custom models) ...

## Wichtiger Hinweis aus Atkins 2012 (supplements):
# Predictions from a Poisson model need to include both fixed and 
# random effects => weil die random intercepts durchs exponieren (=link funct)
# nicht mehr bei 0 sind und dadurch nicht weggelassen werden können (
# population-average (marginal) vs. individual specific (conditional) estimates. 
# diskutiert in Breslow & Clayton 1993. 



library(lme4)
m2.lmer <- lmer(N_dis12 ~  year.c + I(year.c^2) +  
             age.b.70 + num.int.c + 
             (1 + year.c | id), 
           pep, REML=F)
summary(m2.lmer)
plot(m2.lmer)
library(merTools)
# plot results
#newData <- draw(m2.lmer, type = "average")
#newData <- wiggle(newData, varlist = "year.c", 
                  #valueslist = list(unique(pep$year.c)))
# = kürzere Formulierung, aber mit weniger Eingriffsmöglichkeiten
newData <- with(pep, expand.grid(year.c = unique(year.c), 
                                  age.b.70 = mean(age.b.70),
                                 num.int.c = mean(num.int.c),
                                 id = 5))

PI <- predictInterval(merMod = m2.lmer, newdata = newData,
                      level = 0.95, n.sims = 1000, which = "fixed",
                      stat = "median", type="linear.prediction",
                      include.resid.var = F)
newData <- cbind(newData, PI)
library(ggplot2)
ggplot(data = newData, 
       aes(x = year.c, y = fit)) +
  geom_line(aes(y = upr, x = year.c),  
            size = 0.3, alpha=0.5, linetype=2) + 
  geom_line(aes(y = lwr, x = year.c), 
            size = 0.3, alpha=0.5, linetype=2) +
  geom_line(aes(y = fit, x = year.c), size = 1) 




## Vergleich von linear model mit negative binomial
m2.nb <- glmer.nb(N_dis12 ~  year.c + 
                  age.b.70 + female +
                  edu.cont.c + non.white + married + 
                  morbidity.c + obese + num.int.c + 
                  (1 + year.c | id), 
                pep, REML=F)
summary(m2.nb)
plot(m2.nb)
anova(m2.lmer, m2.nb)


describe(pep$year.c)
# negative binomial from GLMMadaptive
library(GLMMadaptive)
m2.poiss.riz <- mixed_model(N_dis12 ~  year.c + age.b.70 + female +
                              edu.cont.c + non.white + married + 
                              morbidity.c + obese + num.int.c, 
                         random = ~ 1 + year.c | id, 
                         family=poisson(), data = pep)
summary(m2.poiss.riz)
anova(m2.lmer, m2.poiss.riz)
exp


fm1 <- mixed_model(fixed = y ~ year * group, random = ~ 1 | id, data = DF,
                   family = binomial())

# catterpillar plot
plotREsim(REsim(m2.lmer, n.sims = 100), stat = 'median', sd = TRUE)

library(merTools)
# plot results
newData <- draw(m2.lmer, type = "average")
newData <- wiggle(newData, varlist = "year.c", 
                  valueslist = list(unique(pep$year.c)))
newData <- wiggle(newData, varlist = "age.b.cat3", 
                  valueslist = list(unique(pep$age.b.cat3)))

PI <- predictInterval(merMod = m2.lmer, newdata = newData,
                      level = 0.95, n.sims = 1000, which = "fixed",
                      stat = "median", type="linear.prediction",
                      include.resid.var = F)
newData <- cbind(newData, PI)
library(ggplot2)
ggplot(data = newData, 
       aes(x = year.c, y = fit)) +
  geom_line(aes(y = upr, x = year.c),  
            size = 0.3, alpha=0.5, linetype=2) + 
  geom_line(aes(y = lwr, x = year.c), 
            size = 0.3, alpha=0.5, linetype=2) +
  geom_line(aes(y = fit, x = year.c), size = 1) +
  facet_wrap(~age.b.cat3)





